---
title: "Pipeline"
author: "Sachin Subedi"
date: "2026-04-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Set Working Directory
```{r}
setwd("~data/")
```
## Data file
```{r}
library(readxl)
library(tidyverse)
library(tidyverse)

seq_data <- read_excel("may.xlsx")
colnames(seq_data)
```

```{r}
library(tibble)

lookup <- tribble(
  ~pattern, ~Family, ~Habitat, ~Trophic_Level, ~Trophic_Niche, ~Primary_Lifestyle, ~Type,

  # --- Seabirds --------------------------------------------------------------
  "murre|auk|guillemot|raxorbill|Razorbill",
      "Alcidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "oystercatcher|Haematopus_ostralegus",
      "Haematopodidae", "Coastal", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "tern|Arctic Tern|Arctic tern|arctic tern|Little_Tern|little_tern",
      "Laridae", "Coastal", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",
  "gull-billed_tern",
      "Laridae", "Coastal", "Carnivore", "Omnivore", "Aerial", "Wild_Bird",
  "Sterna hirundo|Sterna_hirundo|Thalasseus_sandvicensis|sandwich_tern|Sandwich_Tern|Sterna sandvicensis|Sandwich tern|Sandwich Tern|CommonTern|common_tern|Common_tern|Common_Tern|Common tern|common tern|Common Tern|common tern",
      "Laridae", "Marine", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",
  "caspian_tern|Caspian tern",
      "Laridae", "Wetland", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",
  "mediterranean_gull|Mediterranean_gull|Mediterranean gull|Mediterranean Gull",
      "Laridae", "Coastal", "Carnivore", "Omnivore", "Generalist", "Wild_Bird",
  "great_black-baked_gull|Great_Black-Backed_Gull|Great_black-backed_gull|Great_Black_Backed_Gull|Great Black-backed Gull|Gr_bk_bd_gull|black-backed gull|Larus marinus",
      "Laridae", "Coastal", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Larus|gull|Gull|Gull1|Gull10|seagull|Larus_michaellis|Larus michahellis|yellow_leggend-gull|yellow_legged-gull|yellow-legged_gull|Yellow-legged_Gull|yellow-legged-gull|Glaucous_gull|Caspian Gull|Caspian gull",
      "Laridae", "Coastal", "Carnivore", "Omnivore", "Generalist", "Wild_Bird",
  "slender-billed_gull|Little_gull|little_gull|lesser_black-backed_gull|lesser black-backed gull|Lesser black-backed Gull|Lesser Black-backed Gull|Larus_argentatus|Herrin_gull|herring gull|Herring Gull|herring_gull|Herring_gull|Herring_Gull|european_herring_gull|European_herring_gull|European_herring_gullNorway|European herring gull|European Herring Gull|European Herring gull|european herring gull|Larus argentatus|audouins_gull",
      "Laridae", "Coastal", "Carnivore", "Aquatic_Predator", "Generalist", "Wild_Bird",
  "Larus canus|Larus ridibundus|Chroicocephalus ridibundus|Chroicocephalus_ridibundus|black-headed gull|black-headed Gull|Black-headed gull|Black-headed Gull|black-headed_gull|Black-headed_gull|Black-headed_Gull|Black-Headed_Gull|black-headed-gull|black_headed_gull|Black_Headed_Gull|black_headed-gull|black-headead_gull|blach-headed gull|black headed seagull|Bl_H_gull",
      "Laridae", "Coastal", "Carnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Black-legged Kittiwake|Black-legged_kitiwake|black-legged_kittiwake|Black-legged_kittiwake|Kittiwake|kittiwake",
      "Laridae", "Marine", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",

  # --- Cormorants / Pelicans / Penguins -------------------------------------
  "cormorant|Cormorant|Cormorants|great cormorant|Great cormorant|Great Cormorant|Phalacrocorax_carbo",
      "Phalacrocoracidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "Shag",
      "Phalacrocoracidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "dalmatian pelican|spot-billed pelican|zoo_pelican",
      "Pelecanidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "pelican|Pelican",
      "Pelecanidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "penguin",
      "Spheniscidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",

  # --- Marine & semi‑aquatic mammals ----------------------------------------
  "Otter|otter|LutraLutra",
      "Mustelidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "walrus",
      "Odobenidae", "Marine", "Carnivore", "Invertivore", "Aquatic", "Wild_Bird",
  "seal|harbour_seal",
      "Otariidae", "Marine", "Carnivore", "Vertivore", "Aquatic", "Wild_Bird",
  "porpoise",
      "Phocoenidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Mammal",
  "mink|white mink",
      "Mustelidae", "Farm", "Carnivore", "Vertivore", "Generalist", "Domestic_Mammal",

  # --- Waterfowl (Anatidae) --------------------------------------------------
  "Go|goose|Goose|teal|Teal|anser|wild goose|wild_goose|Wild_goose|tundra_bean_goose|Taiga bean goose|Graylag goose|graylag goose|graylag_goose|Grey_Go|Anser anser|bean_goose|Bean goose|bean goose|Eurasian_teal|Eurasian teal|eurasian teal|Common Teal|common teal",
      "Anatidae", "Wetland", "Herbivore", "Herbivore_Terrestrial", "Terrestrial", "Wild_Bird",
  "pomeranian goose|rural_goose|white-fronted Goose|White-fronted Goose|white-fronted goose|white-fronted_goose|White-fronted_goose|Anser albifrons|pink_footed_goose|pink-footed goose|Pink-footed Goose|Pink-footed_goose|pink-footed_goose|Pink-Footed_Goose|Anser brachyrhynchus|Branta ruficollis|Red-breasted_goose|Branta leucopsis|Branta canadensis|Lesser white-fronted goose|lesser white-fronted goose|Greater White-fronted Goose|greater white-fronted goose|greater_white-fronted_goose|Anser albifrons|greater canada goose|Branta bernicla|brant goose|brant|brent_goose|canada goose|Canada goose|Canada Goose|canada_goose|Canada_goose|Canada_Goose|Canadagoose|Canade goose|barnacle goose|Barnacle goose|Barnacle Goose|barnacle_goose|Barnacle_goose|barnacle|Anser_Brachyrhynchus_Anser_Anser",
      "Anatidae", "Grassland", "Herbivore", "Herbivore_Terrestrial", "Terrestrial", "Wild_Bird",
  "Hawaiian_goose|Hawaiian goose|Nene_Goose|Branta sandvicensis",
      "Anatidae", "Shrubland", "Herbivore", "Herbivore_Terrestrial", "Terrestrial", "Wild_Bird",
  "teal|Teal|Eurasian_teal|Eurasian teal|eurasian teal|Common Teal|common teal|Anas crecca",
      "Anatidae", "Wetland", "Herbivore", "Herbivore_Aquatic", "Aquatic", "Wild_Bird",
  "Dk|duck|Duck|wild_duck|wild duck|Anas sp|whistling_duck|Whistling_Duck|mallard_Denmark|Mallard duck|Anas platyrhynchos|anas_platyrhynchos|Mulard duck|Mal|Mallard Duck|mallard_Denmark|mallard_duck|Mallard_duck|mallard|Mallard|cygnus_cygnus|cygnus-cygnus|cygnus_olor|Cygnus_olor|cygnus-olor|cygnus-columbianus|M_Swan|Mute swan|Mute Swan|mute swan|mute_swan|Mute_swan|Mute_Swan|mute-Swan|mute-swan|Muteswan|Cygnus olor|Swan|swan|Whooper swan|Whooper Swan|whooper swan|whooper_swan|Whooper_swan|Whooper_Swan|whooper-swan|Cygnus cygnus|Cygnus olor|Cygnus atratus|black-swan|Black-necked Swan|black-necked swan|black_swan|black swan|Bk_swan|Anas strepera|Gadwall|gadwall|pochard_duck|common_pochard",
      "Anatidae", "Wetland", "Herbivore", "Herbivore_Aquatic", "Aquatic", "Wild_Bird",
  "coscoroba swan",
      "Anatidae", "Wetland", "Omnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "wigeon|Wigeon|Mareca penelope|Anas penelope|eurasian wigeon|Eurasian wigeon|Eur_Wig|Eurasian Wigeon|eurasian_wigeon|Eurasian_wigeon|eurasian-wigeon",
      "Anatidae", "Wetland", "Omnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "Common Goldeneye|common goldeneye|common golden eye|Bucephala clangula",
      "Anatidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "Common eider|common eider|Somateria mollissima|Somateria spectabilis|common_eider|Common_eider|steamer duck",
      "Anatidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "tufted duck|Aythya fuligula|Tufted Duck|Tufted duck|tufted_duck|Aythya_fuligula",
      "Anatidae", "Wetland", "Omnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "shelduck|Tadorna tadorna|common shelduck",
      "Anatidae", "Coastal", "Carnivore", "Aquatic_Predator", "Generalist", "Wild_Bird",
  "Ruddy Shelduck|Tadorna feruginea",
      "Anatidae", "Wetland", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "muscovy duck|Muscovy duck|muscovy_duck|Cairina moschata|mule duck",
      "Anatidae", "Wetland", "Omnivore", "Omnivore", "Generalist", "Wild_Bird",
  "Mandarin Duck",
      "Anatidae", "Forest", "Herbivore", "Herbivore_Aquatic", "Generalist", "Wild_Bird",
  "Carolina_Duck",
      "Anatidae", "Wetland", "Herbivore", "Omnivore", "Generalist", "Wild_Bird",

  # --- Wading birds & shorebirds -------------------------------------------
  "grey heron|Grey heron|Grey Heron|grey_heron|gray_heron|grey-heron|Egretta_garzetta|heron|egret|Ardea_cinerea|Ardea cinerea|Ardea alba|great_egret|great egret|Great Egret",
      "Ardeidae", "Wetland", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "western-cattle-egret",
      "Ardeidae", "Human_Modified", "Carnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "stork|white_stork|White_stork|White Stork|white stork|Ciconia_ciconia|ciconia_ciconia",
      "Ciconiidae", "Grassland", "Carnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "lapwing|northern lapwing",
      "Charadriidae", "Wetland", "Carnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "crane|common_crane|common crane|Common Crane|Common crane|crowned_crane_zoo",
      "Gruidae", "Wetland", "Carnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "flamingo",
      "Phoenicopteridae", "Wetland", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "grebe|Great crested grebe|great crested grebe|Great Crested Grebe|great_crested_grebe|Podiceps_cristatus|Podiceps cristatus|G_c_grebe|Tachybaptus_ruficollis|Tachybaptus ruficollis",
      "Podicipedidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "moorhen|Coot|common coot",
      "Rallidae", "Wetland", "Herbivore", "Herbivore_Aquatic", "Generalist", "Wild_Bird",
  "knot|red knot|Red Knot|Red knot|knot_wader|Calidris canutus|sanderling|Sanderling|Calidris alba|green_sandpiper|Sand_piper",
      "Scolopacidae", "Coastal", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "curlew|Curlew|Eurasian curlew|eurasian curlew|Numenius arquata|Numenius_arquata",
      "Scolopacidae", "Grassland", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "Eurasian Spoonbill|Eurasian spoonbill|eurasian-spoonbill|spoonbill|sacred-ibis",
      "Threskiornithidae", "Wetland", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",

  # --- Humans & marine mammals ---------------------------------------------
  "human",
      "Hominidae", "Urban", "Omnivore", "Omnivore", "Terrestrial", "Human",
  "dolphin",
      "Delphinidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Mammal",

  # --- Tubenoses, skuas, gannets -------------------------------------------
  "Northern Fulmar|northern fulmar|Northern_fulmar|Southern_Fulmar|Fulmar",
      "Procellariidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "skua|parasitic_jaeger|Great skua|Great_skua|great_skua|Great_Skua|long-tailed_skua",
      "Stercorariidae", "Marine", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",
  "Northern gannet|northern gannet|Northern Gannet|northern_gannet|Northern_gannet|Nothern_gannet|nothern_gannet|NothernGannet|gannet|Gannet|Morus_bassanus",
      "Sulidae", "Marine", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",

  # --- Domestic birds & farm animals ---------------------------------------
  "chicken|Ch|Chicken|Poultry|rooster|turkey|layer|hen",
      "Phasianidae", "Farm", "Herbivore", "Omnivore", "Terrestrial", "Domestic_Bird",
  "quail",
      "Phasianidae", "Farm", "Herbivore", "Omnivore", "Terrestrial", "Domestic_Bird",
  "domestic duck|muscovy duck|Muscovy duck|muscovy_duck|Cairina_moschata|Cairina moschata|domestic_duck|Anas platyrhynchos var. domesticus|Anas platyrhynchos f. domestica|Domestic_duck|Domestic_Duck|domestic-duck|pekin_duck|Indian Runner Duck|Indian_runner_ducks|decoy_duck|breeder duck",
      "Anatidae", "Farm", "Herbivore", "Herbivore_Aquatic", "Aquatic", "Domestic_Bird",
  "domestic goose|Anser anser domesticus|domestic_goose|Domestic_goose|Domestic_Goose|captive-goose",
      "Anatidae", "Farm", "Herbivore", "HerbivoreTerrestrial", "Terrestrial", "Domestic_Bird",

  # --- Domestic mammals -----------------------------------------------------
  "domestic|cat",
      "Felidae", "Urban", "Carnivore", "Vertivore", "Terrestrial", "Domestic_Mammal",

  # --- Raptors (Accipitridae) ----------------------------------------------
  "sea_eagle|Sea_eagle|white_tailed_eagle|White_Tailed_Eagle|white-tailed eagle|White-tailed eagle|White-Tailed Eagle|white-tailed_eagle|White-tailed_eagle|white-tailed sea eagle|White-tailed sea eagle|Halietus albicilla",
      "Accipitridae", "Rock", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",
  "Haliaeetus leucocephalus|bald_eagle",
      "Accipitridae", "Wetland", "Carnivore", "Omnivore", "Generalist", "Wild_Bird",
  "golden_eagle",
      "Accipitridae", "Grassland", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",
  "Vulture|vulture|bearded_vulture|griffon vulture",
      "Accipitridae", "Grassland", "Scavenger", "Scavenger", "Terrestrial", "Wild_Bird",
  "bearded_vulture|Gypaetus_barbatus",
      "Accipitridae", "Rock", "Scavenger", "Scavenger", "Terrestrial", "Wild_Bird",
  "buzzard|Buzzard|common buzzard|Common Buzzard|Common buzzard|common_buzzard|Common_buzzard|Common_Buzzard|common-buzzard|rough-legged buzzard|Buteo_buteo|Buteo buteo",
      "Accipitridae", "Grassland", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",
  "goshawk|Goshawk|Eurasian Goshawk|Eurasian goshawk|northern goshawk|Northern goshawk|Northern_goshawk|northern_goshawk|Accipiter gentilis",
      "Accipitridae", "Forest", "Carnivore", "Vertivore", "Generalist", "Wild_Bird",
  "sparrowhawk|Sparrowhawk|eurasian sparrowhawk|Eurasian sparrowhawk|Eurasian Sparrowhawk|Accipiter nisus",
      "Accipitridae", "Forest", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",
  "hawk|red-backed-hawk",
      "Accipitridae", "Woodland", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",
  "Harris Hawk|Harris_Hawk|Parabuteo",
      "Accipitridae", "Shrubland", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",

  # --- Songbirds & others ---------------------------------------------------
  "Reed_Warbler",
      "Acrocephalidae", "Wetland", "Carnivore", "Invertivore", "Insessorial", "Wild_Bird",
  "pigeon|Columba_palumbus|Pigeon",
      "Columbidae", "Woodland", "Herbivore", "Omnivore", "Insessorial", "Wild_Bird",
  "dove|european_turtle_dove",
      "Columbidae", "Woodland", "Herbivore", "Granivore", "Terrestrial", "Wild_Bird",
  "Raven|raven|Common raven|Crow|crow|corvus_corvus",
      "Corvidae", "Forest", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "western jackdaw|jackdaw|Corvus_monedula|pica_pica|magpie|Magpie|common magpie|Eurasian magpie|Eurasian Magpie|Eurasian magpie",
      "Corvidae", "Human_Modified", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Rook|Corvus frugilegus",
      "Corvidae", "Grassland", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "garrulus_glandarius|Garrulus_glandarius",
      "Corvidae", "Forest", "Omnivore", "Omnivore", "Insessorial", "Wild_Bird",

  # --- Falcons --------------------------------------------------------------
  "Falco peregrinus|Falco_peregrinus|falcon|P_falcon|peregrine falcon|Peregrine falcon|Peregrine Falcon|peregrine_Falcon|peregrine_falcon|Peregrine_falcon|Peregrine_Falcon|peregrine-falcon|Perigrine falcon|Pilgrim Falcon|Gyrfalcon|Hobby_Falcon",
      "Falconidae", "Grassland", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",
  "Hobby_Falcon",
      "Falconidae", "Woodland", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",
  "Common Kestrel|common kestrel|common-kestrel|kestrel|Kestrel",
      "Falconidae", "Shrubland", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",

  # --- Other galliforms -----------------------------------------------------
  "Guinea fowl|guinea_fowl|Guinea_Fowl|guinea-fowl|guinea_fowl-chicken|GuineaFowl|guineafowl|Guineafowl|Numida meleagris",
      "Numididae", "Woodland", "Herbivore", "Omnivore", "Terrestrial", "Wild_Bird",

  # --- Passerines & small birds --------------------------------------------
  "house_sparrow",
      "Passeridae", "Human_Modified", "Herbivore", "Granivore", "Terrestrial", "Wild_Bird",

  # --- Pheasants, peafowl, partridge ---------------------------------------
  "pheasant|Pheasant|common pheasant|Common pheasant|Common Pheasant|Phasianus colchicus|Phasianus_colchicus|Phasanius sp.",
      "Phasianidae", "Human_Modified", "Herbivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Indian Peafowl|peacock|Peacock|Common peacock|Pavo cristatus|Pavo|Pavo_cristatus|peafowl|Partridge|partridge",
      "Phasianidae", "Forest", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Partridge|partridge",
      "Phasianidae", "Grassland", "Herbivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Gallus_gallus|Meleagris_gallopavo|Meleagris gallopavo",
      "Phasianidae", "Forest", "Herbivore", "Omnivore", "Terrestrial", "Wild_Bird",

  # --- Ratites --------------------------------------------------------------
  "greater Herbivore_Aquaticrhea|Rhea",
      "Rheidae", "Grassland", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "emy",
      "Casuariidae", "Grassland", "Herbivore", "Omnivore", "Terrestrial", "Wild_Bird",

  # --- Owls -----------------------------------------------------------------
  "Bubo_bubo|Eurasian eagle-owl|Eurasian_eagle-owl|Eurasian eagle owl|Eurasian Eagle-Owl|eagle owl|eagle_owl|Eagle_owl",
      "Strigidae", "Rock", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",
  "Ural owl|ural_owl|Strix uralensis",
      "Strigidae", "Forest", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",
  "little_owl|little-owl",
      "Strigidae", "Human_Modified", "Carnivore", "Omnivore", "Insessorial", "Wild_Bird",
  "long-eared owl",
      "Strigidae", "Forest", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",
  "Otus_scops",
      "Strigidae", "Woodland", "Carnivore", "Invertivore", "Insessorial", "Wild_Bird",
  "tawny owl|Tawny owl|tawny_owl|Towny owel",
      "Strigidae", "Forest", "Carnivore", "Invertivore", "Insessorial", "Wild_Bird",

  # --- Barn owl -------------------------------------------------------------
  "barn owl|Tyto_alba|Tyto alba",
      "Tytonidae", "Human_Modified", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",

  # --- Thrushes & robins ----------------------------------------------------
  "thrush",
      "Turdidae", "Forest", "Omnivore", "Invertivore", "Generalist", "Wild_Bird",
  "blackbird",
      "Turdidae", "Forest", "Omnivore", "Omnivore", "Generalist", "Wild_Bird",

  # --- Robins & kingfishers -------------------------------------------------
  "european_robin",
      "Muscicapidae", "Forest", "Carnivore", "Omnivore", "Generalist", "Wild_Bird",
  "common-kingfisher",
      "Alcedinidae", "Forest", "Carnivore", "Aquatic_Predator", "Insessorial", "Wild_Bird",

  # --- Wild mammals ---------------------------------------------------------
  "fox|Fox|red fox|Red fox|red_fox|Red_fox|Red_Fox|red-fox|Vulpes_vulpes|Vulpes_Vulpes|silver fox|Bush_dog|raccoon_dog|raccoon-dog",
      "Canidae", "Forest", "Carnivore", "Omnivore", "Terrestrial", "Wild_Mammal",
  "lynx|Lynx|caracal|Caracal",
      "Felidae", "Forest", "Carnivore", "Vertivore", "Terrestrial", "Wild_Mammal",
  "European polecat|polecat|marten|ferret|badger|Mustela_putorius|Mustela_furo|badger",
      "Mustelidae", "Forest", "Carnivore", "Vertivore", "Terrestrial", "Wild_Mammal",
  "bear|tibetan blue bear",
      "Ursidae", "Forest", "Carnivore", "Omnivore", "Terrestrial", "Wild_Mammal",

  # --- Catch‑all ------------------------------------------------------------
  "wild bird|Wild bird|wild_bird|avian|Mixed_wild_birds|environment|NA|unknown|Feces|feces|UNL|Wild waterfowl",
      "Others", "Others", "Others", "Others", "Others", "Others"
)

```


```{r}
classify_host <- function(text_value, lookup_table) {
  hits <- lookup_table %>%
    filter(str_detect(text_value, regex(pattern, ignore_case = TRUE)))
  
  if (nrow(hits) == 0) {
    return(rep(NA_character_, 6))
  }
  
  best_match <- hits %>% slice_max(order_by = nchar(pattern), n = 1)
  
  unlist(best_match[c("Family", "Habitat", "Trophic_Level", "Trophic_Niche", "Primary_Lifestyle", "Type")],
         use.names = FALSE)
}
```

## Matched for Isolate_Name
```{r}
seq_data <- seq_data %>%
  mutate(
    matched = map(Isolate_Name, ~ classify_host(.x, lookup)),
    Family            = map_chr(matched, 1),
    Habitat           = map_chr(matched, 2),
    Trophic_Level     = map_chr(matched, 3),
    Trophic_Niche     = map_chr(matched, 4),
    Primary_Lifestyle = map_chr(matched, 5),
    Type              = map_chr(matched, 6)
  ) %>%
  dplyr::select(-matched)


```

## Matched for Host
```{r}
# Apply classification to rows where Family is NA, using Host
seq_data <- seq_data %>%
  mutate(
    matched_fallback = if_else(
      is.na(Family),
      map(Host, ~ classify_host(.x, lookup)),
      list(rep(NA_character_, 6))  # skip if already matched
    ),
    Family = if_else(is.na(Family),        map_chr(matched_fallback, 1), Family),
    Habitat = if_else(is.na(Habitat),      map_chr(matched_fallback, 2), Habitat),
    Trophic_Level = if_else(is.na(Trophic_Level), map_chr(matched_fallback, 3), Trophic_Level),
    Trophic_Niche = if_else(is.na(Trophic_Niche),          map_chr(matched_fallback, 4), Trophic_Niche),
    Primary_Lifestyle = if_else(is.na(Primary_Lifestyle),  map_chr(matched_fallback, 5), Primary_Lifestyle),
    Type = if_else(is.na(Type),            map_chr(matched_fallback, 6), Type)
  ) %>%
  dplyr::select(-matched_fallback)

```

## Remove lab derived, Penguin and Outlier
```{r}
seq_data <- seq_data %>%
  filter(
    !(Host %in% c("Laboratory derived", "Unknown") |
      Family == "Spheniscidae" |
      Isolate_Id == "EPI_ISL_18538407")
  )

```

```{r}
library(stringr)

seq_data$Country <- trimws(str_match(seq_data$Location, "^Europe / ([^/]+)")[, 2])

seq_data <- seq_data %>%
  mutate(Country = case_when(
    Country == "Russian Federation" ~ "Russia",
    Country == "Czech Republic" ~ "Czechia",
    Country == "Macedonia, the former Yogoslav Republic of" ~ "North Macedonia",
    Country == "Moldova, Republic of" ~ "Moldova",
    Country == "Svalbard and Jan Mayen" ~ "Norway",
    Country == "Jersey" ~ "United Kingdom",
    Country == "Isle of Man" ~ "United Kingdom",
    Country == "Bosnia Herz" ~ "Bosnia and Herzegovina",
    Country == "Serbia" ~ "Republic of Serbia",
    TRUE ~ Country
  )) %>%
  filter(Country != "Russia")

cat("Unique countries in seq_data:\n")
print(sort(unique(seq_data$Country)))
```

```{r}
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

unique_countries <- unique(seq_data$Country)

europe_map <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  filter(admin %in% unique_countries)

cat("Matched", nrow(europe_map), "of", length(unique_countries), "countries in map data\n")

centroids <- st_point_on_surface(europe_map) %>% 
             mutate(
               Longitude = st_coordinates(.)[,1],
               Latitude  = st_coordinates(.)[,2]
             ) %>% 
             st_drop_geometry() %>% 
             dplyr::select(Country = admin, Longitude, Latitude)

# check for any countries that rnaturalearth didn't return
anti_join(tibble(Country = unique_countries), centroids, by = "Country")
```


# ============================================================================
# SECTION 7: REGIONAL CLASSIFICATIONS
# ============================================================================


## 7A. EEA Biogeographic Regions (PRIMARY CLASSIFICATION)

Based on European Environment Agency biogeographic regions map.
This addresses Reviewer 2's concern about ecological/biological grounding.

```{r biogeographic-regions}
library(dplyr)
library(sf)
library(rnaturalearth)

region_cols <- c(
  alpine        = "#E69F00",
  atlantic      = "#CC79A7",
  boreal        = "#56B4E9",
  continental   = "#009E73",
  mediterranean = "#0072B2",
  pannonian     = "#D55E00",
  steppic       = "#000000"
)

biogeo_sf <- st_read("~/data/BioGeo/BiogeoRegions2016.shp", quiet = TRUE) %>%
  st_make_valid() %>%
  filter(short_name %in% names(region_cols)) %>%
  select(EEA_Bioregion_raw = short_name, geometry)

eu_countries <- ne_countries(continent = "Europe", returnclass = "sf") %>%
  st_make_valid()

eu_no_russia <- eu_countries %>%
  filter(admin != "Russia") %>%
  st_transform(st_crs(biogeo_sf))

eu_mask <- st_union(eu_no_russia)
eu_mask_sf <- st_sf(id = 1, geometry = eu_mask)

biogeo_no_russia <- st_intersection(biogeo_sf, eu_mask_sf)

study_countries <- eu_no_russia %>%
  rename(Country = admin) %>%
  filter(Country %in% unique(seq_data$Country))

study_countries <- study_countries %>%
  mutate(
    Country = case_when(
      Country == "Bosnia and Herz." ~ "Bosnia and Herzegovina",
      Country == "Serbia" ~ "Republic of Serbia",
      TRUE ~ Country
    )
  )

study_countries_ea <- st_transform(study_countries, 3035)
biogeo_no_russia_ea <- st_transform(biogeo_no_russia, 3035)

country_region <- st_intersection(
  study_countries_ea %>% select(Country),
  biogeo_no_russia_ea %>% select(EEA_Bioregion_raw)
)

country_region_weights <- country_region %>%
  mutate(overlap_area = st_area(geometry)) %>%
  st_drop_geometry() %>%
  group_by(Country, EEA_Bioregion_raw) %>%
  summarise(overlap_area = sum(overlap_area), .groups = "drop") %>%
  left_join(
    study_countries_ea %>%
      mutate(country_area = st_area(geometry)) %>%
      st_drop_geometry() %>%
      select(Country, country_area),
    by = "Country"
  ) %>%
  mutate(weight = as.numeric(overlap_area / country_area)) %>%
  mutate(
    EEA_Bioregion = case_when(
      EEA_Bioregion_raw == "alpine" ~ "Alpine",
      EEA_Bioregion_raw == "atlantic" ~ "Atlantic",
      EEA_Bioregion_raw == "boreal" ~ "Boreal",
      EEA_Bioregion_raw == "continental" ~ "Continental",
      EEA_Bioregion_raw == "mediterranean" ~ "Mediterranean",
      EEA_Bioregion_raw == "pannonian" ~ "Pannonian",
      EEA_Bioregion_raw == "steppic" ~ "Steppic",
      TRUE ~ EEA_Bioregion_raw
    )
  ) %>%
  arrange(Country, desc(weight))

country_bioregion_primary <- country_region_weights %>%
  group_by(Country) %>%
  slice_max(order_by = weight, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    Country,
    EEA_Bioregion,
    p_primary = weight
  )

island_bio <- data.frame(
  Country       = c("Iceland", "Cyprus"),
  EEA_Bioregion = c("Atlantic", "Mediterranean"),
  p_primary     = c(1.0, 1.0),
  stringsAsFactors = FALSE
)
country_bioregion_primary <- bind_rows(
  country_bioregion_primary %>% dplyr::filter(!Country %in% island_bio$Country),
  island_bio
)

country_bioregion_primary <- country_bioregion_primary %>%
  mutate(
    Bioregion_5 = case_when(
      EEA_Bioregion %in% c("Boreal", "Arctic") ~ "Northern",
      EEA_Bioregion == "Atlantic" ~ "Atlantic",
      EEA_Bioregion %in% c("Continental", "Pannonian", "Steppic") ~ "Continental",
      EEA_Bioregion == "Mediterranean" ~ "Mediterranean",
      EEA_Bioregion == "Alpine" ~ "Alpine",
      TRUE ~ NA_character_
    )
  )

cat("\nCountry → Dominant Bioregion (area-weighted):\n")
print(as.data.frame(country_bioregion_primary), row.names = FALSE)

seq_data <- seq_data %>%
  dplyr::select(-any_of(c("EEA_Bioregion", "Bioregion_5", "p_primary"))) %>%
  left_join(country_bioregion_primary, by = "Country")

missing_bio <- seq_data %>% filter(is.na(EEA_Bioregion)) %>% distinct(Country)
if (nrow(missing_bio) > 0) {
  cat("Countries missing biogeographic assignment:\n")
  print(missing_bio)
}

cat("\nEEA Biogeographic Region counts:\n")
print(table(seq_data$EEA_Bioregion, useNA = "ifany"))

cat("\nSimplified 5-Region counts:\n")
print(table(seq_data$Bioregion_5, useNA = "ifany"))

cat("\nPrimary-region overlap proportion summary (p_primary):\n")
print(summary(seq_data$p_primary))
```
## 7B2. Hierarchical Clustering — Mainland Europe

```{r hierarchical-clustering-no-cyprus}
set.seed(21)
library(dplyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(RColorBrewer)
library(geosphere)
library(cluster)
library(dendextend)

island_outliers <- c("Iceland", "Cyprus")

centroids_main <- centroids %>% filter(!Country %in% island_outliers)

coords_main <- as.matrix(centroids_main[, c("Longitude", "Latitude")])
rownames(coords_main) <- centroids_main$Country

D_main <- distm(coords_main, fun = distHaversine) / 1000   # km
rownames(D_main) <- centroids_main$Country
colnames(D_main) <- centroids_main$Country
D_main_dist <- as.dist(D_main)

linkage_methods <- c("ward.D2", "complete", "average")

coph_main <- sapply(linkage_methods, function(m) {
  hc <- hclust(D_main_dist, method = m)
  cor(D_main_dist, cophenetic(hc))
})

cat("Cophenetic correlations (mainland, excl. Cyprus):\n")
print(round(coph_main, 3))

best_method_main <- names(which.max(coph_main))
cat("\nSelected linkage method (mainland):", best_method_main, "\n")

hc_main <- hclust(D_main_dist, method = best_method_main)

k_range <- 3:10

sil_main <- sapply(k_range, function(k) {
  cl <- cutree(hc_main, k = k)
  mean(silhouette(cl, D_main_dist)[, "sil_width"])
})

sil_main_df <- data.frame(k = k_range, silhouette = round(sil_main, 3))
cat("\nSilhouette scores by k (mainland only, excl. Cyprus):\n")
print(sil_main_df)

k_opt_main <- k_range[which.max(sil_main)]
cat("\nOptimal k (mainland):", k_opt_main, "\n")

cl_main <- cutree(hc_main, k = k_opt_main)

serbia_cluster <- paste0("HC_Cluster_", cl_main[["Republic of Serbia"]])
uk_cluster     <- paste0("HC_Cluster_", cl_main[["United Kingdom"]])
greece_cluster <- paste0("HC_Cluster_", cl_main[["Greece"]])

hc_lookup_initial <- data.frame(
  Country    = c(names(cl_main), "Cyprus", "Belarus", "Montenegro", "Iceland"),
  HC_Cluster = c(
    paste0("HC_Cluster_", cl_main),
    greece_cluster,
    serbia_cluster,
    serbia_cluster,
    uk_cluster
  ),
  stringsAsFactors = FALSE
)

cat("\nHierarchical Cluster composition — initial (mainland + manual islands):\n")
cat("(Cyprus → Greece cluster:", greece_cluster, ")\n")
cat("(Belarus & Montenegro → Serbia cluster:", serbia_cluster, ")\n")
cat("(Iceland → UK cluster:", uk_cluster, ")\n")
for (i in sort(unique(hc_lookup_initial$HC_Cluster))) {
  m <- hc_lookup_initial$Country[hc_lookup_initial$HC_Cluster == i]
  cat(sprintf("\n%s (n=%d): %s\n", i, length(m), paste(m, collapse = ", ")))
}

hc_map_initial <- hc_lookup_initial %>% rename(HC_Cluster_main = HC_Cluster)

europe_hc_initial <- ne_countries(continent = "Europe", returnclass = "sf") %>%
  filter(name != "Russia", admin != "Iceland") %>%
  left_join(hc_map_initial, by = c("admin" = "Country"))

n_ini <- length(unique(na.omit(europe_hc_initial$HC_Cluster_main)))
pal_ini <- setNames(
  brewer.pal(max(3, n_ini), "Set2")[seq_len(n_ini)],
  sort(unique(na.omit(europe_hc_initial$HC_Cluster_main)))
)

p_hc_initial <- ggplot(europe_hc_initial) +
  geom_sf(aes(fill = HC_Cluster_main), colour = "black", linewidth = 0.2) +
  scale_fill_manual(values = pal_ini, na.value = "gray90",
                    name = "HC Cluster\n(initial)") +
  coord_sf(xlim = c(-25, 45), ylim = c(34, 72), expand = FALSE) +
  labs(title = paste0("HC — initial (k = ", k_opt_main, ", tree + island rules)")) +
  theme_minimal() +
  theme(legend.position = "right", axis.text = element_blank(),
        axis.title = element_blank(), panel.grid = element_blank())

print(p_hc_initial)

ggsave("HC_Mainland_Map_initial.png", p_hc_initial,
       width = 10, height = 8, dpi = 300, bg = "white")
ggsave("HC_Mainland_Map_initial.pdf", p_hc_initial,
       width = 10, height = 8, device = cairo_pdf)

hc_lookup_main <- hc_lookup_initial %>%
  mutate(
    HC_Cluster = case_when(
      Country == "Norway" & HC_Cluster == "HC_Cluster_5" ~ "HC_Cluster_3",
      TRUE ~ HC_Cluster
    )
  )

cat("\n--- After Norway recode (HC_Cluster_5 → HC_Cluster_3) ---\n")
print(hc_lookup_main %>% filter(Country == "Norway"))

centroids <- centroids %>%
  dplyr::select(-any_of("HC_Cluster_main")) %>%
  left_join(hc_lookup_main %>% rename(HC_Cluster_main = HC_Cluster), by = "Country")

seq_data <- seq_data %>%
  dplyr::select(-any_of("HC_Cluster_main")) %>%
  left_join(hc_lookup_main %>% rename(HC_Cluster_main = HC_Cluster), by = "Country")

cat("\nHC_Cluster_main counts in seq_data (after recode):\n")
print(table(seq_data$HC_Cluster_main, useNA = "ifany"))

dend_main <- as.dendrogram(hc_main)
dend_main <- color_branches(dend_main, k = k_opt_main)
dend_main <- color_labels(dend_main, k = k_opt_main)

par(mar = c(12, 4, 4, 2))
plot(dend_main,
     main = paste0("Hierarchical Clustering — Mainland Europe", k_opt_main, ")"),
     ylab = "Haversine Distance (km)", cex = 0.8)
rect.dendrogram(dend_main, k = k_opt_main,
                border = 2:(k_opt_main + 1), lty = 2)

png("Dendrogram_HC_mainland.png", width = 14, height = 8, units = "in", res = 300)
par(mar = c(12, 4, 4, 2))
plot(dend_main,
     main = paste0("HC Mainland excl. Cyprus (k = ", k_opt_main, ")"),
     ylab = "Haversine Distance (km)", cex = 0.8)
rect.dendrogram(dend_main, k = k_opt_main,
                border = 2:(k_opt_main + 1), lty = 2)
dev.off()

hc_main_lookup_for_map <- hc_lookup_main %>% rename(HC_Cluster_main = HC_Cluster)

europe_hc_main <- ne_countries(continent = "Europe", returnclass = "sf") %>%
  filter(name != "Russia", admin != "Iceland") %>%
  left_join(hc_main_lookup_for_map, by = c("admin" = "Country"))

n_hc_main   <- length(unique(na.omit(europe_hc_main$HC_Cluster_main)))
hc_main_pal <- setNames(
  brewer.pal(max(3, n_hc_main), "Set2")[seq_len(n_hc_main)],
  sort(unique(na.omit(europe_hc_main$HC_Cluster_main)))
)

p_hc_main <- ggplot(europe_hc_main) +
  geom_sf(aes(fill = HC_Cluster_main), colour = "black", linewidth = 0.2) +
  scale_fill_manual(values = hc_main_pal, na.value = "gray90",
                    name = "HC Cluster\n(recode)") +
  coord_sf(xlim = c(-25, 45), ylim = c(34, 72), expand = FALSE) +
  labs(title = paste0("HC — Norway HC_Cluster_5 → HC_Cluster_3 (k_tree = ",
                      k_opt_main, ")")) +
  theme_minimal() +
  theme(legend.position = "right", axis.text = element_blank(),
        axis.title = element_blank(), panel.grid = element_blank())

print(p_hc_main)

ggsave("HC_Mainland_Map.png", p_hc_main,
       width = 10, height = 8, dpi = 300, bg = "white")
ggsave("HC_Mainland_Map.pdf", p_hc_main,
       width = 10, height = 8, device = cairo_pdf)
```


## 7C. Combined Hierarchical Clustering and EEA biogeographic regions

```{r combined-hc-biogeography}
library(dplyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(RColorBrewer)

seq_data <- seq_data %>%
  dplyr::select(-any_of(c("HC_Bio_Combined", "HC_Bio_ID", "HC_Bio5_Combined"))) %>%
  mutate(
    HC_Bio_Combined = case_when(
      is.na(HC_Cluster_main) | is.na(EEA_Bioregion) ~ NA_character_,
      TRUE ~ paste0(HC_Cluster_main, "_", EEA_Bioregion)
    ),
    HC_Bio_ID = case_when(
      is.na(HC_Cluster_main) | is.na(EEA_Bioregion) ~ NA_character_,
      TRUE ~ paste0(HC_Cluster_main, "_", EEA_Bioregion)
    ),
    HC_Bio5_Combined = case_when(
      is.na(HC_Cluster_main) | is.na(Bioregion_5) ~ NA_character_,
      TRUE ~ paste0(HC_Cluster_main, "_", Bioregion_5)
    )
  )

eu_country_key <- function(admin_chr) {
  case_when(
    admin_chr %in% c("Macedonia", "Macedonia, the former Yugoslav Republic of") ~ "North Macedonia",
    admin_chr == "Bosnia and Herz." ~ "Bosnia and Herzegovina",
    admin_chr == "Serbia" ~ "Republic of Serbia",
    TRUE ~ as.character(admin_chr)
  )
}

hc_bio_lookup_for_map <- seq_data %>%
  mutate(Country = eu_country_key(Country)) %>%
  filter(!is.na(HC_Bio_ID)) %>%
  distinct(Country, HC_Bio_ID, HC_Bio_Combined)

serbia_hc <- hc_lookup_main %>%
  filter(Country == "Republic of Serbia") %>%
  slice(1) %>%
  pull(HC_Cluster)

belarus_hc <- hc_lookup_main %>%
  filter(Country == "Belarus") %>%
  slice(1) %>%
  pull(HC_Cluster)

montenegro_bio <- country_bioregion_primary %>%
  filter(Country == "Montenegro") %>%
  slice(1) %>%
  pull(EEA_Bioregion)

if (length(montenegro_bio) == 0 || is.na(montenegro_bio)) {
  montenegro_bio <- "Alpine"
}

manual_hc_bio_map <- tibble(
  Country = c("Belarus", "Montenegro"),
  HC_Bio_ID = c(
    paste(belarus_hc, "Continental", sep = "_"),
    paste(serbia_hc, montenegro_bio, sep = "_")
  ),
  HC_Bio_Combined = c(
    paste0(belarus_hc, "_", "Continental"),
    paste0(serbia_hc, "_", montenegro_bio)
  )
)

hc_bio_lookup_for_map <- bind_rows(
  hc_bio_lookup_for_map,
  anti_join(manual_hc_bio_map, hc_bio_lookup_for_map, by = "Country")
)

europe_hc_bio <- ne_countries(continent = "Europe", returnclass = "sf") %>%
  filter(name != "Russia", admin != "Iceland") %>%
  mutate(Country = eu_country_key(admin)) %>%
  left_join(hc_bio_lookup_for_map, by = "Country")

cat("\nMontenegro row on map:\n")
print(
  europe_hc_bio %>%
    st_drop_geometry() %>%
    filter(Country == "Montenegro") %>%
    select(admin, Country, HC_Bio_ID, HC_Bio_Combined)
)

hc_bio_levels <- sort(unique(na.omit(europe_hc_bio$HC_Bio_Combined)))
europe_hc_bio$HC_Bio_Combined <- factor(europe_hc_bio$HC_Bio_Combined, levels = hc_bio_levels)

n_hc_bio <- length(hc_bio_levels)

hc_cols <- c(
  HC_Cluster_1_Alpine = "#CC79A7",
  HC_Cluster_1_Arctic = "#8DA0CB",
  HC_Cluster_1_Atlantic = "#CD5C5C",
  HC_Cluster_1_Continental = "#BCBD22",
  HC_Cluster_2_Alpine = "#26A69A",
  HC_Cluster_2_Continental = "#0072B2",
  HC_Cluster_2_Mediterranean = "lightslategray",
  HC_Cluster_2_Pannonian = "#6B4C1B",
  HC_Cluster_3_Alpine = "#7570B3",
  HC_Cluster_3_Boreal = "#006D2C",
  HC_Cluster_4_Mediterranean = "#56B4E9"
)

default_hc_bio_col <- "#BDBDBD"
hc_bio_pal <- stats::setNames(
  vapply(hc_bio_levels, function(lvl) {
    if (lvl %in% names(hc_cols)) unname(hc_cols[[lvl]]) else default_hc_bio_col
  }, character(1)),
  hc_bio_levels
)

p_hc_bio <- ggplot(europe_hc_bio) +
  geom_sf(aes(fill = HC_Bio_Combined), colour = "black", linewidth = 0.2) +
  scale_fill_manual(
    values = hc_bio_pal,
    na.value = "gray90",
    name = "HC and EEA bioregion\n(HC_Bio_Combined)",
    drop = FALSE
  ) +
  coord_sf(xlim = c(-25, 45), ylim = c(34, 72), expand = FALSE) +
  labs(
    title = "Combined hierarchical cluster and dominant EEA bioregion",
    subtitle = paste0(n_hc_bio, " unique HC_bioregion strata in study countries")
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = rel(0.78)),
    legend.key.size = grid::unit(0.45, "cm"),
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

print(p_hc_bio)

ggsave("HC_Bio_ID_Map.png", p_hc_bio,
       width = 10, height = 8, dpi = 300, bg = "white")

ggsave("HC_Bio_ID_Map.pdf", p_hc_bio,
       width = 10, height = 8, device = cairo_pdf)
```


```{r}
p_hc_bio <- ggplot(europe_hc_bio) +
  geom_sf(aes(fill = HC_Bio_Combined), colour = "black", linewidth = 0.2) +
  scale_fill_manual(
    values = hc_bio_pal,
    na.value = "gray90",
    name = "HC and EEA bioregion\n(HC_Bio_Combined)",
    drop = FALSE
  ) +
  coord_sf(xlim = c(-25, 45), ylim = c(34, 72), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

print(p_hc_bio)

ggsave("HC_Bio_ID_Map_no_legend.png", p_hc_bio,
       width = 10, height = 8, dpi = 300, bg = "white")

ggsave("HC_Bio_ID_Map_no_legend.pdf", p_hc_bio,
       width = 10, height = 8, device = cairo_pdf)

ggsave("HC_Bio_ID_Map_no_legend_transparent.png", p_hc_bio,
       width = 10, height = 8, dpi = 300, bg = "transparent")
```


```{r}
library(dplyr)

# Define the Flyway classification mapping
flyway_mapping <- c(
  "Denmark" = "EA",
  "Germany" = "EA",
  "Poland" = "EA",
  "Netherlands" = "EA",
  "Sweden" = "EA",
  "France" = "EA",
  "United Kingdom" = "EA",
  "Belgium" = "EA",
  "Spain" = "EA",
  "Ireland" = "EA",
  "Norway" = "EA",
  "Finland" = "EA",
  "Latvia" = "EA",
  "Estonia" = "EA",
  "Lithuania" = "EA",
  "Luxembourg" = "EA",
  "Iceland" = "EA",
  "Portugal" = "EA",
  "Switzerland" = "MBS",
  "Ukraine" = "MBS",
  "Hungary" = "MBS",
  "Croatia" = "MBS",
  "Italy" = "MBS",
  "North Macedonia" = "MBS",
  "Czechia" = "MBS",
  "Greece" = "MBS",
  "Bulgaria" = "MBS",
  "Slovakia" = "MBS",
  "Slovenia" = "MBS",
  "Romania" = "MBS",
  "Austria" = "MBS",
  "Serbia" = "MBS",
  "Republic of Serbia" = "MBS",
  "Montenegro" = "MBS",
  "Belarus" = "MBS",
  "Kosovo" = "MBS",
  "Albania" = "MBS",
  "Bosnia and Herzegovina" = "MBS",
  "Moldova" = "MBS",
  "Russia" = "WAEA",
  "Cyprus" = "WAEA"
)

# Add the new Flyway column to the dataset
seq_data <- seq_data %>%
  mutate(Flyway = flyway_mapping[Country])

```


```{r}
library(dplyr)
library(stringr)
library(lubridate)

# Helper to normalize join keys
normalize_id <- function(x) {
  x %>%
    as.character() %>%
    str_trim() %>%
    str_squish() %>%
    toupper()
}

# Make sure Collection_Date is character
seq_data <- seq_data %>%
  mutate(
    Isolate_Id = as.character(Isolate_Id),
    Collection_Date = as.character(Collection_Date)
  ) %>%
  select(-any_of(c(
    "Collection_Date_Fixed",
    "Collection_Date_Fixed_chr",
    "Collection_Date_Parsed",
    "Collection_Date_Final",
    "Date_Imputed"
  )))

# 1) Identify incomplete dates
incomplete_dates <- seq_data %>%
  filter(!is.na(Collection_Date)) %>%
  filter(
    grepl("^\\d{4}$", Collection_Date) |
    grepl("^\\d{4}-\\d{2}$", Collection_Date)
  )

message("Incomplete original dates: ", nrow(incomplete_dates))

# 2) Create fixed dates
incomplete_dates <- incomplete_dates %>%
  mutate(
    Collection_Date_Fixed_chr = case_when(
      grepl("^\\d{4}$", Collection_Date) ~ paste0(Collection_Date, "-07-01"),
      grepl("^\\d{4}-\\d{2}$", Collection_Date) ~ paste0(Collection_Date, "-15"),
      TRUE ~ Collection_Date
    ),
    Collection_Date_Fixed = suppressWarnings(ymd(Collection_Date_Fixed_chr))
  )

# Optional export
incomplete_dates %>%
  select(Isolate_Id, Collection_Date, Collection_Date_Fixed) %>%
  write.table(
    "incomplete_dates_summary.txt",
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )

# 3) Normalize keys
seq_norm <- seq_data %>%
  mutate(Isolate_Id = normalize_id(Isolate_Id))

inc_norm <- incomplete_dates %>%
  mutate(Isolate_Id = normalize_id(Isolate_Id)) %>%
  arrange(Isolate_Id, desc(!is.na(Collection_Date_Fixed))) %>%
  group_by(Isolate_Id) %>%
  summarise(
    Collection_Date_Fixed = first(Collection_Date_Fixed[!is.na(Collection_Date_Fixed)], default = as.Date(NA)),
    .groups = "drop"
  )

# Diagnostics
overlap <- inner_join(
  seq_norm %>% distinct(Isolate_Id),
  inc_norm %>% distinct(Isolate_Id),
  by = "Isolate_Id"
) %>% nrow()

message("Key overlap after normalization: ", overlap)

# 4) Join + create final date + flag
seq_data <- seq_norm %>%
  left_join(inc_norm, by = "Isolate_Id") %>%
  mutate(
    Collection_Date_Parsed = suppressWarnings(ymd(Collection_Date)),
    Collection_Date_Final = coalesce(Collection_Date_Fixed, Collection_Date_Parsed),
    Date_Imputed = !is.na(Collection_Date_Fixed)
  ) %>%
  select(-Collection_Date_Fixed)

# 5) Sanity checks
print(table(seq_data$Date_Imputed, useNA = "ifany"))
cat("Remaining NA final dates:", sum(is.na(seq_data$Collection_Date_Final)), "\n")
```


```{r}
## Traits Summary
library(dplyr)
library(lubridate)

seq_data <- seq_data %>%
  mutate(Year = year(as.Date(Collection_Date_Final)))

# ---- Summary counts for each trait ----
family_counts <- seq_data %>%
  group_by(Family) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

habitat_counts <- seq_data %>%
  group_by(Habitat) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

trophic_level_counts <- seq_data %>%
  group_by(Trophic_Level) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

trophic_niche_counts <- seq_data %>%
  group_by(Trophic_Niche) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

lifestyle_counts <- seq_data %>%
  group_by(Primary_Lifestyle) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

type_counts <- seq_data %>%
  group_by(Type) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

flyway_counts <- seq_data %>%
  group_by(Flyway) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

HC_Cluster_main_counts <- seq_data %>%
  group_by(HC_Cluster_main) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

country_counts <- seq_data %>%
  group_by(Country) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

# ---- Yearly isolate summary ----
yearly_counts <- seq_data %>%
  group_by(Year) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Year))

# ---- Region-level isolate summary ----
EEA_Bioregion_counts <- seq_data %>%
  group_by(EEA_Bioregion) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

HC_Bio_combined_counts <- seq_data %>%
  group_by(HC_Bio_Combined) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

HC_Bio5_combined_counts <- seq_data %>%
  group_by(HC_Bio5_Combined) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))


# ---- Print summaries ----
print(family_counts)
print(habitat_counts)
print(trophic_level_counts)
print(trophic_niche_counts)
print(lifestyle_counts)
print(type_counts)
print(country_counts)
print(flyway_counts)
print(yearly_counts)
print(HC_Cluster_main_counts)
print(EEA_Bioregion_counts)
print(HC_Bio_combined_counts)
print(HC_Bio5_combined_counts)
```

```{r}
library(dplyr)

final_data <- seq_data %>%
  mutate(
    Isolate_Name = gsub(" ", "_", Isolate_Name),
    Country = gsub(" ", "_", Country)
  ) %>%
  dplyr::select(
    Isolate_Id, Isolate_Name, Habitat, Type, Country, Flyway,
    HC_Cluster_main, EEA_Bioregion,
    HC_Bio_Combined, Collection_Date_Final, Sequence
  )

# Save the filtered dataset as a TSV
write.table(final_data, "final_seq_data_updated.tsv", sep="\t", row.names=FALSE, quote=FALSE)
```

