---
title: "Pipeline for Data Cleaning"
author: "Sachin Subedi"
date: "2025-12-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Set Working Directory
```{r}
setwd("~/Library/CloudStorage/OneDrive-UniversityofGeorgia/EU_H5/may/data/Again/after_draft/pipeline/GitHub/")
```

## Metadata file from GISAID
```{r}
library(readxl)
library(tidyverse)
library(tidyverse)

seq_data <- read_excel("may.xlsx")
colnames(seq_data)
```

## Lookup file for those wild bird species I have from GISAID data and categorize them according to traits from AVONET 
```{r}
library(tibble)

lookup <- tribble(
  ~pattern, ~Family, ~Habitat, ~Trophic_Level, ~Trophic_Niche, ~Primary_Lifestyle, ~Type,

  # --- Seabirds --------------------------------------------------------------
  "murre|auk|guillemot|raxorbill|Razorbill",
      "Alcidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "oystercatcher|Haematopus_ostralegus",
      "Haematopodidae", "Coastal", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "tern|Arctic Tern|Arctic tern|arctic tern|Little_Tern|little_tern",
      "Laridae", "Coastal", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",
  "gull-billed_tern",
      "Laridae", "Coastal", "Carnivore", "Omnivore", "Aerial", "Wild_Bird",
  "Sterna hirundo|Sterna_hirundo|Thalasseus_sandvicensis|sandwich_tern|Sandwich_Tern|Sterna sandvicensis|Sandwich tern|Sandwich Tern|CommonTern|common_tern|Common_tern|Common_Tern|Common tern|common tern|Common Tern|common tern",
      "Laridae", "Marine", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",
  "caspian_tern|Caspian tern",
      "Laridae", "Wetland", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",
  "mediterranean_gull|Mediterranean_gull|Mediterranean gull|Mediterranean Gull",
      "Laridae", "Coastal", "Carnivore", "Omnivore", "Generalist", "Wild_Bird",
  "great_black-baked_gull|Great_Black-Backed_Gull|Great_black-backed_gull|Great_Black_Backed_Gull|Great Black-backed Gull|Gr_bk_bd_gull|black-backed gull|Larus marinus",
      "Laridae", "Coastal", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Larus|gull|Gull|Gull1|Gull10|seagull|Larus_michaellis|Larus michahellis|yellow_leggend-gull|yellow_legged-gull|yellow-legged_gull|Yellow-legged_Gull|yellow-legged-gull|Glaucous_gull|Caspian Gull|Caspian gull",
      "Laridae", "Coastal", "Carnivore", "Omnivore", "Generalist", "Wild_Bird",
  "slender-billed_gull|Little_gull|little_gull|lesser_black-backed_gull|lesser black-backed gull|Lesser black-backed Gull|Lesser Black-backed Gull|Larus_argentatus|Herrin_gull|herring gull|Herring Gull|herring_gull|Herring_gull|Herring_Gull|european_herring_gull|European_herring_gull|European_herring_gullNorway|European herring gull|European Herring Gull|European Herring gull|european herring gull|Larus argentatus|audouins_gull",
      "Laridae", "Coastal", "Carnivore", "Aquatic_Predator", "Generalist", "Wild_Bird",
  "Larus canus|Larus ridibundus|Chroicocephalus ridibundus|Chroicocephalus_ridibundus|black-headed gull|black-headed Gull|Black-headed gull|Black-headed Gull|black-headed_gull|Black-headed_gull|Black-headed_Gull|Black-Headed_Gull|black-headed-gull|black_headed_gull|Black_Headed_Gull|black_headed-gull|black-headead_gull|blach-headed gull|black headed seagull|Bl_H_gull",
      "Laridae", "Coastal", "Carnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Black-legged Kittiwake|Black-legged_kitiwake|black-legged_kittiwake|Black-legged_kittiwake|Kittiwake|kittiwake",
      "Laridae", "Marine", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",

  # --- Cormorants / Pelicans / Penguins -------------------------------------
  "cormorant|Cormorant|Cormorants|great cormorant|Great cormorant|Great Cormorant|Phalacrocorax_carbo",
      "Phalacrocoracidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "Shag",
      "Phalacrocoracidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "dalmatian pelican|spot-billed pelican|zoo_pelican",
      "Pelecanidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "pelican|Pelican",
      "Pelecanidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "penguin",
      "Spheniscidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",

  # --- Marine & semi‑aquatic mammals ----------------------------------------
  "Otter|otter|LutraLutra",
      "Mustelidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "walrus",
      "Odobenidae", "Marine", "Carnivore", "Invertivore", "Aquatic", "Wild_Bird",
  "seal|harbour_seal",
      "Otariidae", "Marine", "Carnivore", "Vertivore", "Aquatic", "Wild_Bird",
  "porpoise",
      "Phocoenidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Mammal",
  "mink|white mink",
      "Mustelidae", "Farm", "Carnivore", "Vertivore", "Generalist", "Domestic_Mammal",

  # --- Waterfowl (Anatidae) --------------------------------------------------
  "Go|goose|Goose|teal|Teal|anser|wild goose|wild_goose|Wild_goose|tundra_bean_goose|Taiga bean goose|Graylag goose|graylag goose|graylag_goose|Grey_Go|Anser anser|bean_goose|Bean goose|bean goose|Eurasian_teal|Eurasian teal|eurasian teal|Common Teal|common teal",
      "Anatidae", "Wetland", "Herbivore", "Herbivore_Terrestrial", "Terrestrial", "Wild_Bird",
  "pomeranian goose|rural_goose|white-fronted Goose|White-fronted Goose|white-fronted goose|white-fronted_goose|White-fronted_goose|Anser albifrons|pink_footed_goose|pink-footed goose|Pink-footed Goose|Pink-footed_goose|pink-footed_goose|Pink-Footed_Goose|Anser brachyrhynchus|Branta ruficollis|Red-breasted_goose|Branta leucopsis|Branta canadensis|Lesser white-fronted goose|lesser white-fronted goose|Greater White-fronted Goose|greater white-fronted goose|greater_white-fronted_goose|Anser albifrons|greater canada goose|Branta bernicla|brant goose|brant|brent_goose|canada goose|Canada goose|Canada Goose|canada_goose|Canada_goose|Canada_Goose|Canadagoose|Canade goose|barnacle goose|Barnacle goose|Barnacle Goose|barnacle_goose|Barnacle_goose|barnacle|Anser_Brachyrhynchus_Anser_Anser",
      "Anatidae", "Grassland", "Herbivore", "Herbivore_Terrestrial", "Terrestrial", "Wild_Bird",
  "Hawaiian_goose|Hawaiian goose|Nene_Goose|Branta sandvicensis",
      "Anatidae", "Shrubland", "Herbivore", "Herbivore_Terrestrial", "Terrestrial", "Wild_Bird",
  "teal|Teal|Eurasian_teal|Eurasian teal|eurasian teal|Common Teal|common teal|Anas crecca",
      "Anatidae", "Wetland", "Herbivore", "Herbivore_Aquatic", "Aquatic", "Wild_Bird",
  "Dk|duck|Duck|wild_duck|wild duck|Anas sp|whistling_duck|Whistling_Duck|mallard_Denmark|Mallard duck|Anas platyrhynchos|anas_platyrhynchos|Mulard duck|Mal|Mallard Duck|mallard_Denmark|mallard_duck|Mallard_duck|mallard|Mallard|cygnus_cygnus|cygnus-cygnus|cygnus_olor|Cygnus_olor|cygnus-olor|cygnus-columbianus|M_Swan|Mute swan|Mute Swan|mute swan|mute_swan|Mute_swan|Mute_Swan|mute-Swan|mute-swan|Muteswan|Cygnus olor|Swan|swan|Whooper swan|Whooper Swan|whooper swan|whooper_swan|Whooper_swan|Whooper_Swan|whooper-swan|Cygnus cygnus|Cygnus olor|Cygnus atratus|black-swan|Black-necked Swan|black-necked swan|black_swan|black swan|Bk_swan|Anas strepera|Gadwall|gadwall|pochard_duck|common_pochard",
      "Anatidae", "Wetland", "Herbivore", "Herbivore_Aquatic", "Aquatic", "Wild_Bird",
  "coscoroba swan",
      "Anatidae", "Wetland", "Omnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "wigeon|Wigeon|Mareca penelope|Anas penelope|eurasian wigeon|Eurasian wigeon|Eur_Wig|Eurasian Wigeon|eurasian_wigeon|Eurasian_wigeon|eurasian-wigeon",
      "Anatidae", "Wetland", "Omnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "Common Goldeneye|common goldeneye|common golden eye|Bucephala clangula",
      "Anatidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "Common eider|common eider|Somateria mollissima|Somateria spectabilis|common_eider|Common_eider|steamer duck",
      "Anatidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "tufted duck|Aythya fuligula|Tufted Duck|Tufted duck|tufted_duck|Aythya_fuligula",
      "Anatidae", "Wetland", "Omnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "shelduck|Tadorna tadorna|common shelduck",
      "Anatidae", "Coastal", "Carnivore", "Aquatic_Predator", "Generalist", "Wild_Bird",
  "Ruddy Shelduck|Tadorna feruginea",
      "Anatidae", "Wetland", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "muscovy duck|Muscovy duck|muscovy_duck|Cairina moschata|mule duck",
      "Anatidae", "Wetland", "Omnivore", "Omnivore", "Generalist", "Wild_Bird",
  "Mandarin Duck",
      "Anatidae", "Forest", "Herbivore", "Herbivore_Aquatic", "Generalist", "Wild_Bird",
  "Carolina_Duck",
      "Anatidae", "Wetland", "Herbivore", "Omnivore", "Generalist", "Wild_Bird",

  # --- Wading birds & shorebirds -------------------------------------------
  "grey heron|Grey heron|Grey Heron|grey_heron|gray_heron|grey-heron|Egretta_garzetta|heron|egret|Ardea_cinerea|Ardea cinerea|Ardea alba|great_egret|great egret|Great Egret",
      "Ardeidae", "Wetland", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "western-cattle-egret",
      "Ardeidae", "Human_Modified", "Carnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "stork|white_stork|White_stork|White Stork|white stork|Ciconia_ciconia|ciconia_ciconia",
      "Ciconiidae", "Grassland", "Carnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "lapwing|northern lapwing",
      "Charadriidae", "Wetland", "Carnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "crane|common_crane|common crane|Common Crane|Common crane|crowned_crane_zoo",
      "Gruidae", "Wetland", "Carnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "flamingo",
      "Phoenicopteridae", "Wetland", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "grebe|Great crested grebe|great crested grebe|Great Crested Grebe|great_crested_grebe|Podiceps_cristatus|Podiceps cristatus|G_c_grebe|Tachybaptus_ruficollis|Tachybaptus ruficollis",
      "Podicipedidae", "Wetland", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "moorhen|Coot|common coot",
      "Rallidae", "Wetland", "Herbivore", "Herbivore_Aquatic", "Generalist", "Wild_Bird",
  "knot|red knot|Red Knot|Red knot|knot_wader|Calidris canutus|sanderling|Sanderling|Calidris alba|green_sandpiper|Sand_piper",
      "Scolopacidae", "Coastal", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "curlew|Curlew|Eurasian curlew|eurasian curlew|Numenius arquata|Numenius_arquata",
      "Scolopacidae", "Grassland", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",
  "Eurasian Spoonbill|Eurasian spoonbill|eurasian-spoonbill|spoonbill|sacred-ibis",
      "Threskiornithidae", "Wetland", "Carnivore", "Aquatic_Predator", "Terrestrial", "Wild_Bird",

  # --- Humans & marine mammals ---------------------------------------------
  "human",
      "Hominidae", "Urban", "Omnivore", "Omnivore", "Terrestrial", "Human",
  "dolphin",
      "Delphinidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Mammal",

  # --- Tubenoses, skuas, gannets -------------------------------------------
  "Northern Fulmar|northern fulmar|Northern_fulmar|Southern_Fulmar|Fulmar",
      "Procellariidae", "Marine", "Carnivore", "Aquatic_Predator", "Aquatic", "Wild_Bird",
  "skua|parasitic_jaeger|Great skua|Great_skua|great_skua|Great_Skua|long-tailed_skua",
      "Stercorariidae", "Marine", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",
  "Northern gannet|northern gannet|Northern Gannet|northern_gannet|Northern_gannet|Nothern_gannet|nothern_gannet|NothernGannet|gannet|Gannet|Morus_bassanus",
      "Sulidae", "Marine", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",

  # --- Domestic birds & farm animals ---------------------------------------
  "chicken|Ch|Chicken|Poultry|rooster|turkey|layer|hen",
      "Phasianidae", "Farm", "Herbivore", "Omnivore", "Terrestrial", "Domestic_Bird",
  "quail",
      "Phasianidae", "Farm", "Herbivore", "Omnivore", "Terrestrial", "Domestic_Bird",
  "domestic duck|muscovy duck|Muscovy duck|muscovy_duck|Cairina_moschata|Cairina moschata|domestic_duck|Anas platyrhynchos var. domesticus|Anas platyrhynchos f. domestica|Domestic_duck|Domestic_Duck|domestic-duck|pekin_duck|Indian Runner Duck|Indian_runner_ducks|decoy_duck|breeder duck",
      "Anatidae", "Farm", "Herbivore", "Herbivore_Aquatic", "Aquatic", "Domestic_Bird",
  "domestic goose|Anser anser domesticus|domestic_goose|Domestic_goose|Domestic_Goose|captive-goose",
      "Anatidae", "Farm", "Herbivore", "HerbivoreTerrestrial", "Terrestrial", "Domestic_Bird",

  # --- Domestic mammals -----------------------------------------------------
  "domestic|cat",
      "Felidae", "Urban", "Carnivore", "Vertivore", "Terrestrial", "Domestic_Mammal",

  # --- Raptors (Accipitridae) ----------------------------------------------
  "sea_eagle|Sea_eagle|white_tailed_eagle|White_Tailed_Eagle|white-tailed eagle|White-tailed eagle|White-Tailed Eagle|white-tailed_eagle|White-tailed_eagle|white-tailed sea eagle|White-tailed sea eagle|Halietus albicilla",
      "Accipitridae", "Rock", "Carnivore", "Aquatic_Predator", "Aerial", "Wild_Bird",
  "Haliaeetus leucocephalus|bald_eagle",
      "Accipitridae", "Wetland", "Carnivore", "Omnivore", "Generalist", "Wild_Bird",
  "golden_eagle",
      "Accipitridae", "Grassland", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",
  "Vulture|vulture|bearded_vulture|griffon vulture",
      "Accipitridae", "Grassland", "Scavenger", "Scavenger", "Terrestrial", "Wild_Bird",
  "bearded_vulture|Gypaetus_barbatus",
      "Accipitridae", "Rock", "Scavenger", "Scavenger", "Terrestrial", "Wild_Bird",
  "buzzard|Buzzard|common buzzard|Common Buzzard|Common buzzard|common_buzzard|Common_buzzard|Common_Buzzard|common-buzzard|rough-legged buzzard|Buteo_buteo|Buteo buteo",
      "Accipitridae", "Grassland", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",
  "goshawk|Goshawk|Eurasian Goshawk|Eurasian goshawk|northern goshawk|Northern goshawk|Northern_goshawk|northern_goshawk|Accipiter gentilis",
      "Accipitridae", "Forest", "Carnivore", "Vertivore", "Generalist", "Wild_Bird",
  "sparrowhawk|Sparrowhawk|eurasian sparrowhawk|Eurasian sparrowhawk|Eurasian Sparrowhawk|Accipiter nisus",
      "Accipitridae", "Forest", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",
  "hawk|red-backed-hawk",
      "Accipitridae", "Woodland", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",
  "Harris Hawk|Harris_Hawk|Parabuteo",
      "Accipitridae", "Shrubland", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",

  # --- Songbirds & others ---------------------------------------------------
  "Reed_Warbler",
      "Acrocephalidae", "Wetland", "Carnivore", "Invertivore", "Insessorial", "Wild_Bird",
  "pigeon|Columba_palumbus|Pigeon",
      "Columbidae", "Woodland", "Herbivore", "Omnivore", "Insessorial", "Wild_Bird",
  "dove|european_turtle_dove",
      "Columbidae", "Woodland", "Herbivore", "Granivore", "Terrestrial", "Wild_Bird",
  "Raven|raven|Common raven|Crow|crow|corvus_corvus",
      "Corvidae", "Forest", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "western jackdaw|jackdaw|Corvus_monedula|pica_pica|magpie|Magpie|common magpie|Eurasian magpie|Eurasian Magpie|Eurasian magpie",
      "Corvidae", "Human_Modified", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Rook|Corvus frugilegus",
      "Corvidae", "Grassland", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "garrulus_glandarius|Garrulus_glandarius",
      "Corvidae", "Forest", "Omnivore", "Omnivore", "Insessorial", "Wild_Bird",

  # --- Falcons --------------------------------------------------------------
  "Falco peregrinus|Falco_peregrinus|falcon|P_falcon|peregrine falcon|Peregrine falcon|Peregrine Falcon|peregrine_Falcon|peregrine_falcon|Peregrine_falcon|Peregrine_Falcon|peregrine-falcon|Perigrine falcon|Pilgrim Falcon|Gyrfalcon|Hobby_Falcon",
      "Falconidae", "Grassland", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",
  "Hobby_Falcon",
      "Falconidae", "Woodland", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",
  "Common Kestrel|common kestrel|common-kestrel|kestrel|Kestrel",
      "Falconidae", "Shrubland", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",

  # --- Other galliforms -----------------------------------------------------
  "Guinea fowl|guinea_fowl|Guinea_Fowl|guinea-fowl|guinea_fowl-chicken|GuineaFowl|guineafowl|Guineafowl|Numida meleagris",
      "Numididae", "Woodland", "Herbivore", "Omnivore", "Terrestrial", "Wild_Bird",

  # --- Passerines & small birds --------------------------------------------
  "house_sparrow",
      "Passeridae", "Human_Modified", "Herbivore", "Granivore", "Terrestrial", "Wild_Bird",

  # --- Pheasants, peafowl, partridge ---------------------------------------
  "pheasant|Pheasant|common pheasant|Common pheasant|Common Pheasant|Phasianus colchicus|Phasianus_colchicus|Phasanius sp.",
      "Phasianidae", "Human_Modified", "Herbivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Indian Peafowl|peacock|Peacock|Common peacock|Pavo cristatus|Pavo|Pavo_cristatus|peafowl|Partridge|partridge",
      "Phasianidae", "Forest", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Partridge|partridge",
      "Phasianidae", "Grassland", "Herbivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "Gallus_gallus|Meleagris_gallopavo|Meleagris gallopavo",
      "Phasianidae", "Forest", "Herbivore", "Omnivore", "Terrestrial", "Wild_Bird",

  # --- Ratites --------------------------------------------------------------
  "greater Herbivore_Aquaticrhea|Rhea",
      "Rheidae", "Grassland", "Omnivore", "Omnivore", "Terrestrial", "Wild_Bird",
  "emy",
      "Casuariidae", "Grassland", "Herbivore", "Omnivore", "Terrestrial", "Wild_Bird",

  # --- Owls -----------------------------------------------------------------
  "Bubo_bubo|Eurasian eagle-owl|Eurasian_eagle-owl|Eurasian eagle owl|Eurasian Eagle-Owl|eagle owl|eagle_owl|Eagle_owl",
      "Strigidae", "Rock", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",
  "Ural owl|ural_owl|Strix uralensis",
      "Strigidae", "Forest", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",
  "little_owl|little-owl",
      "Strigidae", "Human_Modified", "Carnivore", "Omnivore", "Insessorial", "Wild_Bird",
  "long-eared owl",
      "Strigidae", "Forest", "Carnivore", "Vertivore", "Aerial", "Wild_Bird",
  "Otus_scops",
      "Strigidae", "Woodland", "Carnivore", "Invertivore", "Insessorial", "Wild_Bird",
  "tawny owl|Tawny owl|tawny_owl|Towny owel",
      "Strigidae", "Forest", "Carnivore", "Invertivore", "Insessorial", "Wild_Bird",

  # --- Barn owl -------------------------------------------------------------
  "barn owl|Tyto_alba|Tyto alba",
      "Tytonidae", "Human_Modified", "Carnivore", "Vertivore", "Insessorial", "Wild_Bird",

  # --- Thrushes & robins ----------------------------------------------------
  "thrush",
      "Turdidae", "Forest", "Omnivore", "Invertivore", "Generalist", "Wild_Bird",
  "blackbird",
      "Turdidae", "Forest", "Omnivore", "Omnivore", "Generalist", "Wild_Bird",

  # --- Robins & kingfishers -------------------------------------------------
  "european_robin",
      "Muscicapidae", "Forest", "Carnivore", "Omnivore", "Generalist", "Wild_Bird",
  "common-kingfisher",
      "Alcedinidae", "Forest", "Carnivore", "Aquatic_Predator", "Insessorial", "Wild_Bird",

  # --- Wild mammals ---------------------------------------------------------
  "fox|Fox|red fox|Red fox|red_fox|Red_fox|Red_Fox|red-fox|Vulpes_vulpes|Vulpes_Vulpes|silver fox|Bush_dog|raccoon_dog|raccoon-dog",
      "Canidae", "Forest", "Carnivore", "Omnivore", "Terrestrial", "Wild_Mammal",
  "lynx|Lynx|caracal|Caracal",
      "Felidae", "Forest", "Carnivore", "Vertivore", "Terrestrial", "Wild_Mammal",
  "European polecat|polecat|marten|ferret|badger|Mustela_putorius|Mustela_furo|badger",
      "Mustelidae", "Forest", "Carnivore", "Vertivore", "Terrestrial", "Wild_Mammal",
  "bear|tibetan blue bear",
      "Ursidae", "Forest", "Carnivore", "Omnivore", "Terrestrial", "Wild_Mammal",

  # --- Catch‑all ------------------------------------------------------------
  "wild bird|Wild bird|wild_bird|avian|Mixed_wild_birds|environment|NA|unknown|Feces|feces|UNL|Wild waterfowl",
      "Others", "Others", "Others", "Others", "Others", "Others"
)

```

## Matching up the wild bird species
```{r}
classify_host <- function(text_value, lookup_table) {
  hits <- lookup_table %>%
    filter(str_detect(text_value, regex(pattern, ignore_case = TRUE)))
  
  if (nrow(hits) == 0) {
    return(rep(NA_character_, 6))
  }
  
  best_match <- hits %>% slice_max(order_by = nchar(pattern), n = 1)
  
  unlist(best_match[c("Family", "Habitat", "Trophic_Level", "Trophic_Niche", "Primary_Lifestyle", "Type")],
         use.names = FALSE)
}
```



## Matched for Isolate_Name
```{r}
seq_data <- seq_data %>%
  mutate(
    matched = map(Isolate_Name, ~ classify_host(.x, lookup)),
    Family            = map_chr(matched, 1),
    Habitat           = map_chr(matched, 2),
    Trophic_Level     = map_chr(matched, 3),
    Trophic_Niche     = map_chr(matched, 4),
    Primary_Lifestyle = map_chr(matched, 5),
    Type              = map_chr(matched, 6)
  ) %>%
  dplyr::select(-matched)


```

## Matched for Host
```{r}
seq_data <- seq_data %>%
  mutate(
    matched_fallback = if_else(
      is.na(Family),
      map(Host, ~ classify_host(.x, lookup)),
      list(rep(NA_character_, 6))  # skip if already matched
    ),
    Family = if_else(is.na(Family),        map_chr(matched_fallback, 1), Family),
    Habitat = if_else(is.na(Habitat),      map_chr(matched_fallback, 2), Habitat),
    Trophic_Level = if_else(is.na(Trophic_Level), map_chr(matched_fallback, 3), Trophic_Level),
    Trophic_Niche = if_else(is.na(Trophic_Niche),          map_chr(matched_fallback, 4), Trophic_Niche),
    Primary_Lifestyle = if_else(is.na(Primary_Lifestyle),  map_chr(matched_fallback, 5), Primary_Lifestyle),
    Type = if_else(is.na(Type),            map_chr(matched_fallback, 6), Type)
  ) %>%
  dplyr::select(-matched_fallback)

```

## Remove lab derived, Penguin and Outlier
```{r}
seq_data <- seq_data %>%
  filter(
    !(Host %in% c("Laboratory derived", "Unknown") |
      Family == "Spheniscidae" |
      Isolate_Id == "EPI_ISL_18538407")
  )

```

## Cleaning the country
```{r}
seq_data$Country <- sub("Europe / ([^/]+)(?: /.*)?", "\\1", seq_data$Location)
```


## Renaming some countries and removing Russia
```{r}
seq_data <- seq_data %>%
  mutate(Country = case_when(
    Country == "Russian Federation" ~ "Russia",
    Country == "Czech Republic" ~ "Czechia",
    Country == "Macedonia, the former Yogoslav Republic of" ~ "North Macedonia",
    Country == "Moldova, Republic of" ~ "Moldova",
    Country == "Svalbard and Jan Mayen" ~ "Norway",
    Country == "Jersey" ~ "United Kingdom",
    Country == "Isle of Man" ~ "United Kingdom",
    Country == "Bosnia Herz" ~ "Bosnia and Herzegovina",
    Country == "Serbia" ~ "Republic of Serbia",
    TRUE ~ Country
  )) %>%
  filter(Country != "Russia")
```

## Getting latitude and longitude coordinates for all the countries
```{r}
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

unique_countries <- unique(seq_data$Country)

europe_map <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  filter(admin %in% unique_countries)


centroids <- st_centroid(europe_map, of_largest_polygon = TRUE) %>% 
             mutate(
               Longitude = st_coordinates(.)[,1],
               Latitude  = st_coordinates(.)[,2]
             ) %>% 
             st_drop_geometry() %>% 
             dplyr::select(Country = admin, Longitude, Latitude)

# check for any countries that rnaturalearth didn't return
anti_join(tibble(Country = unique_countries), centroids, by = "Country")
```

## K-mean Clustering of countries to get GeoCluster
```{r}
library(dplyr)
library(sf)
library(geosphere)
library(ggplot2)
library(ggrepel)
library(rnaturalearth)
library(RColorBrewer)

set.seed(21)

k_init  <- 20
min_n   <- 3

km <- kmeans(centroids[, c("Longitude", "Latitude")],
             centers = k_init, nstart = 100)
centroids$Cluster_raw <- km$cluster

nearest_big <- function(id){
  big <- as.integer(names(which(table(centroids$Cluster_raw) >= min_n)))
  from <- km$centers[id , , drop = FALSE]
  to   <- km$centers[big, , drop = FALSE]
  big[ which.min(distHaversine(from, to)) ]
}

small_ids <- as.integer(names(which(table(centroids$Cluster_raw) < min_n)))
for(s in small_ids){
  centroids$Cluster_raw[centroids$Cluster_raw == s] <- nearest_big(s)
}

ordinals   <- c("One","Two","Three","Four","Five","Six","Seven","Eight","Nine","Ten",
                "Eleven","Twelve","Thirteen","Fourteen","Fifteen","Sixteen","Seventeen",
                "Eighteen","Nineteen","Twenty")

final_ids  <- sort(unique(centroids$Cluster_raw))
n_final    <- length(final_ids)

centroids <- centroids %>%
  mutate(Cluster_num = match(Cluster_raw, final_ids),
         Cluster     = paste0("GeoCluster_", ordinals[Cluster_num])) %>%
  dplyr::select(-Cluster_num)

centroids$Cluster <- factor(
  centroids$Cluster,
  levels = paste0("GeoCluster_", ordinals[seq_len(n_final)])
)

cat("Final cluster sizes:\n")
print(table(centroids$Cluster))

pal <- brewer.pal(max(3, n_final), "Set2")[seq_len(n_final)]
names(pal) <- levels(centroids$Cluster)

europe_sub <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(admin %in% centroids$Country)

cluster_plot <- ggplot() +
  geom_sf(data = europe_sub, fill = "gray95", colour = "gray60") +
  geom_point(data = centroids,
             aes(Longitude, Latitude, colour = Cluster), size = 3.5) +
  geom_text_repel(data = centroids,
                  aes(Longitude, Latitude, label = Country, colour = Cluster),
                  size = 4, max.overlaps = Inf, force = 1.5) +
  scale_colour_manual(values = pal, name = "GeoCluster") +
  coord_sf(xlim = c(-15, 40), ylim = c(35, 71)) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right",
        axis.title = element_blank(),
        axis.text  = element_blank(),
        panel.grid = element_blank())

print(cluster_plot)
ggsave("GeoCluster_Map.png", cluster_plot, width = 10, height = 7,
       dpi = 300, bg = "transparent")

lookup_cty_cluster <- centroids %>%
  dplyr::select(Country, GeoCluster = Cluster)

write.csv(lookup_cty_cluster, "cluster_lookup_auto.csv", row.names = FALSE)

seq_data <- seq_data %>%
  left_join(lookup_cty_cluster, by = "Country")

if(anyNA(seq_data$GeoCluster)){
  warning("Some isolates lack a cluster label – check country spelling:")
  print(seq_data %>% filter(is.na(GeoCluster)) %>% distinct(Country))
}

```



## Working on incomplete dates
```{r}
library(dplyr)
library(stringr)
library(lubridate)

incomplete_dates <- seq_data %>%
  filter(grepl("^\\d{4}$", Collection_Date) | grepl("^\\d{4}-\\d{2}$", Collection_Date))

print("Incomplete original dates:")
print(incomplete_dates)

incomplete_dates <- incomplete_dates %>%
  mutate(Collection_Date_Fixed = case_when(
    grepl("^\\d{4}$", Collection_Date) ~ paste0(Collection_Date, "-07-01"),
    grepl("^\\d{4}-\\d{2}$", Collection_Date) ~ paste0(Collection_Date, "-15"),
    TRUE ~ Collection_Date
  ))

print("Corrected dates:")
print(incomplete_dates %>% dplyr::select(Isolate_Id, Collection_Date, Collection_Date_Fixed))

export_df <- incomplete_dates %>%
  dplyr::select(Isolate_Id, Collection_Date, Collection_Date_Fixed)

write.table(
  export_df,
  file = "incomplete_dates_summary.txt",
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

seq_data$Isolate_Id <- as.character(seq_data$Isolate_Id)
incomplete_dates$Isolate_Id <- as.character(incomplete_dates$Isolate_Id)

seq_data <- seq_data %>%
  left_join(
    incomplete_dates %>% dplyr::select(Isolate_Id, Collection_Date_Fixed),
    by = "Isolate_Id"
  )

seq_data <- seq_data %>%
  mutate(Date_Imputed = !is.na(Collection_Date_Fixed))


normalize_id <- function(x){
  x %>%
    as.character() %>%
    str_trim() %>%
    str_squish() %>%
    toupper()
}

seq_data <- seq_data %>%
  mutate(
    Collection_Date = as.character(Collection_Date),
    Isolate_Id      = normalize_id(Isolate_Id)
  )

incomplete_dates <- seq_data %>%
  filter(!is.na(Collection_Date)) %>%
  filter(
    grepl("^\\d{4}$", Collection_Date) |
    grepl("^\\d{4}-\\d{2}$", Collection_Date)
  )

message("Incomplete original dates: ", nrow(incomplete_dates))

incomplete_dates <- incomplete_dates %>%
  mutate(
    Collection_Date_Fixed_chr = case_when(
      grepl("^\\d{4}$", Collection_Date)        ~ paste0(Collection_Date, "-07-01"),
      grepl("^\\d{4}-\\d{2}$", Collection_Date) ~ paste0(Collection_Date, "-15"),
      TRUE                                      ~ Collection_Date
    ),
    Collection_Date_Fixed = suppressWarnings(ymd(Collection_Date_Fixed_chr))
  )

incomplete_dates %>%
  dplyr::select(Isolate_Id, Collection_Date, Collection_Date_Fixed) %>%
  write.table("incomplete_dates_summary.txt",
              sep = "\t", row.names = FALSE, quote = FALSE)

inc_norm <- incomplete_dates %>%
  mutate(Isolate_Id = normalize_id(Isolate_Id)) %>%
  arrange(Isolate_Id, desc(!is.na(Collection_Date_Fixed))) %>%
  group_by(Isolate_Id) %>%
  summarise(
    Collection_Date_Imputed = first(na.omit(Collection_Date_Fixed)),
    .groups = "drop"
  )

overlap <- inner_join(
  seq_data %>% distinct(Isolate_Id),
  inc_norm  %>% distinct(Isolate_Id),
  by = "Isolate_Id"
) %>% nrow()
message("Key overlap after normalization: ", overlap)

seq_data <- seq_data %>%
  left_join(inc_norm, by = "Isolate_Id") %>%
  mutate(
    Collection_Date_Parsed = suppressWarnings(ymd(Collection_Date)),
    Collection_Date_Final  = coalesce(Collection_Date_Imputed, Collection_Date_Parsed),
    Date_Imputed           = !is.na(Collection_Date_Imputed)
  ) %>%
  dplyr::select(-Collection_Date_Imputed)

print(table(seq_data$Date_Imputed, useNA = "ifany"))
cat("Remaining NA final dates:", sum(is.na(seq_data$Collection_Date_Final)), "\n")

```
## Traits Summary
```{r}
library(dplyr)
library(lubridate)

seq_data <- seq_data %>%
  mutate(Year = year(as.Date(Collection_Date_Final)))

family_counts <- seq_data %>%
  group_by(Family) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

habitat_counts <- seq_data %>%
  group_by(Habitat) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

trophic_level_counts <- seq_data %>%
  group_by(Trophic_Level) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

trophic_niche_counts <- seq_data %>%
  group_by(Trophic_Niche) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

lifestyle_counts <- seq_data %>%
  group_by(Primary_Lifestyle) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

type_counts <- seq_data %>%
  group_by(Type) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

country_counts <- seq_data %>%
  group_by(Country) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))

yearly_counts <- seq_data %>%
  group_by(Year) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Year))

GeoCluster_counts <- seq_data %>%
  group_by(GeoCluster) %>%
  summarise(Isolate_Count = n(), .groups = "drop") %>%
  arrange(desc(Isolate_Count))


print(family_counts)
print(habitat_counts)
print(trophic_level_counts)
print(trophic_niche_counts)
print(lifestyle_counts)
print(type_counts)
print(country_counts)
print(yearly_counts)
print(GeoCluster_counts)
```

## Exporting file in .tsv
```{r}
library(dplyr)

seq_data <- seq_data %>%
  mutate(Isolate_Name = gsub(" ", "_", Isolate_Name))
seq_data <- seq_data %>%
  mutate(Country = gsub(" ", "_", Country))

final_data <- seq_data %>% dplyr::select(Isolate_Id, Isolate_Name, Habitat, Type, Country, GeoCluster, Collection_Date_Final, Sequence)
write.table(final_data, "final_seq_data.tsv", sep="\t", row.names=FALSE, quote=FALSE)
```

## K-means clustering diagnostic
```{r}
library(geosphere)
library(cluster)
library(clusterSim)
library(fpc)
library(purrr)
library(ggplot2)
library(factoextra)
library(RColorBrewer)



set.seed(21)

merged_labels <- function(k, coords, min_n = 3, seed = 21){
  set.seed(seed)
  km <- kmeans(coords, centers = k, nstart = 100)
  cl <- km$cluster
  
  nearest_big <- function(id, cl_vec){
    big <- as.integer(names(which(table(cl_vec) >= min_n)))
    from <- km$centers[id , , drop = FALSE]
    to   <- km$centers[big, , drop = FALSE]
    big[ which.min(distHaversine(from, to)) ]
  }
  repeat{
    small <- as.integer(names(which(table(cl) < min_n)))
    if(length(small) == 0) break
    for(s in small) cl[cl == s] <- nearest_big(s, cl)
  }
  cl
}

coords   <- as.matrix(centroids[, c("Longitude","Latitude")])
D_gc     <- distm(coords) / 1000
k_range  <- 2:10

labels_list <- map(k_range, merged_labels, coords = coords, min_n = 3)

idx_df <- map_df(seq_along(k_range), \(i){
  k  <- k_range[i]
  cl <- labels_list[[i]]
  sil <- silhouette(cl, D_gc)
  
  wss <- sum(apply(coords, 2, function(v) tapply(v, cl, var) * (table(cl) - 1)))
  
  tibble(
    k,
    wss     = wss,
    sil_avg = mean(sil[, "sil_width"]),
    ch      = calinhara(coords, cl),
    db      = index.DB(coords, cl)$DB
  )
})

gap <- clusGap(coords, FUN = kmeans,
               K.max = max(k_range), B = 500, nstart = 50, verbose = FALSE)

k_now   <- n_final
cl_now  <- labels_list[[which(k_range == k_now)]]
sil_now <- silhouette(cl_now, D_gc)


p_elbow <- ggplot(idx_df, aes(k, wss)) +
  geom_line(size = 1) + geom_point(size = 2) +
  scale_x_continuous(breaks = k_range) +
  labs(title = "Elbow plot", y = "Total within-cluster SS", x = "k") +
  theme_minimal(base_size = 13)

p_sil_avg <- ggplot(idx_df, aes(k, sil_avg)) +
  geom_line(size = 1) + geom_point(size = 2) +
  scale_x_continuous(breaks = k_range) +
  labs(title = "Average silhouette width", y = "Silhouette", x = "k") +
  theme_minimal(base_size = 13)

p_ch_db <- ggplot(idx_df, aes(k)) +
  geom_line(aes(y = ch / max(ch)), colour = "steelblue", size = 1) +
  geom_point(aes(y = ch / max(ch)), colour = "steelblue", size = 2) +
  geom_line(aes(y = db / max(db)), colour = "firebrick", size = 1) +
  geom_point(aes(y = db / max(db)), colour = "firebrick", size = 2) +
  scale_x_continuous(breaks = k_range) +
  scale_y_continuous(sec.axis = dup_axis(name = "Scaled DB")) +
  labs(title = "Calinski–Harabasz Index CH (blue ↑)\nvs Davies–Bouldin Index DB (red ↓)", y = "Scaled CH", x = "k") +
  theme_minimal(base_size = 13)

p_gap <- fviz_gap_stat(gap) +
  ggtitle("Gap statistic (raw k-means)")

p_sil_detail <- fviz_silhouette(
  sil_now,
  palette = brewer.pal(k_now, "Set2"),
  ggtheme  = theme_minimal(base_size = 13),
  print.summary = FALSE
) +
  ggtitle(paste0("Silhouette plot (k = ", k_now, ", post-merge)"))


stab <- clusterboot(coords,
                    B = 100, seed = 21,
                    clustermethod = kmeansCBI,
                    krange = k_now,
                    bootmethod = "boot")

cat("\nPer-cluster Jaccard similarities (≥ 0.75 = stable):\n")
print(stab$bootmean)

dir.create("Figures", showWarnings = FALSE)

plots <- list(
  FigS1a_Elbow        = p_elbow,
  FigS1b_SilhouetteAvg= p_sil_avg,
  FigS1c_CH_vs_DB     = p_ch_db,
  FigS1d_Gap                    = p_gap,
  FigS1e_SilhouetteDetail       = p_sil_detail
)

for(nm in names(plots)){
  ggsave(file.path("Figures", paste0(nm, ".png")),
         plots[[nm]], width = 9, height = 5, dpi = 300, bg = "white")
  ggsave(file.path("Figures", paste0(nm, ".pdf")),
         plots[[nm]], width = 9, height = 5, device = cairo_pdf)
}

write.csv(idx_df,
          "Table_Sx_internal_indices.csv", row.names = FALSE)
write.csv(gap$Tab,
          "Table_Sx_gap_statistic.csv", row.names = FALSE)
write.csv(data.frame(Cluster = levels(centroids$Cluster),
                     Jaccard = round(stab$bootmean, 3)),
          "Table_Sx_bootstrap_Jaccard.csv", row.names = FALSE)

cat("\nPost-merge diagnostic figures saved to 'Figures/' and tables written.\n")

```

